Thermal reactionome of a temperate ant species

### 2014-05-20 doc/ApTranscriptome_manuscript.Rmd
Edits to ms

---
title: "Thermal reactionome of a temperate ant species"
author: "John Stanton-Geddes, Andrew Nguyen, Mahesh Vangala, Jim Vincent, Rob Dunn,
  Aaron Ellison, Nate Sanders, Nick Gotelli, Sara Helms-Cahan"
output:
  pdf_document:
    fig_caption: yes
    latex_engine: xelatex
  html_document: null
bibliography: ApTranscriptome.bib
csl: ecology.csl
---

@@

When ants are exposed to different temperatures, lots of genes change expression.

@@

While characterizing the reactionome is useful for description of the molecular pathways for thermal tolerance, a greater understanding of the process of evolution can be gained using a comparative approach [@harvey1991]. Specifically, comparisons of populations or species in contrasting environments can reveal the extent to which geographic differentiation and selection has changed the number of genes involved in thermal tolerance. To accomplish this, we compared the thermal reactionomes between the two species of *Aphaenogaster*. *A. picea*, the northern species, experiences less chronic and acute heat stress than *A. carolensis*. Thus, we predicted that *A. picea* would have more genes than *A. carolinensis* that are up-regulated to high temperatures, while *A. carolinensis* would have higher expression of these thermally-tolerant genes near optimal temperatures. Reciprocally, we expected *A. carolinensis* to have greater up-regulation of genes in response to cold temperatures, while *A. picea* would have greater basal expression of cold-tolerant genes. Finally, phenotypic studies in *Aphaenogaster* [@warren2013] and other ecotherms [@hoffmann2013] have found less variation in critical maximum than minimum temperatures. From this, we hypothesized that if the species had more similar maximum temperatures than minimum temperatures, the northern species *A. picea* would have a broader thermal breadth of gene expression than *A. carolinensis*.

@@

To describe the overall molecular toolkit for thermal tolerance in *Aphaenogaster*, we performed gene set enrichment analysis (GSEA) on the intersection of each expression category in the two species using the program `topGO` [@alexa2006; @alexa2010] with the `parentChild` algorithm [@grossmann2007] implemented in R [@r_core_team2013]. Briefly, this approach identifies GO terms that are overrepresented in the significant genes relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms. To reduce redundancy in the enriched GO terms, we used the web program [ReviGO](http://revigo.irb.hr/).

@@

We identified `r prettyNum(length(which(RxNpval$pval < 0.05)), big.mark = ",", scientific = FALSE)` genes with overall model fits of *P* < 0.05, retaining `r prettyNum(length(which(RxNpval$pval < 0.05)), big.mark = ",", scientific = FALSE)` after adjusting *P* values for false discovery rate. Of these genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` significantly change expression with temperature including `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` that have a species x temperature interaction. The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not thermally-responsive in one species (*NotResp*) differed significantly between the species ($\chi^2$ = `r chi1$statistic`, P < `r max(chi1$p.value, 0.001)`; Table 1). For both species, the majority of responsive transcripts were in the *Low* category (Fig. 1). More transcripts had *Intermediate* expression in *A. carolinensis* than in *A. picea*, which had nearly twice as many bimodally-expressed transcripts  (Fig. 1, Table 1).

@@

We performed a number of additional comparisons to evaluate the extent to which the thermal reactionomes differed between *A. carolinensis* and *A. picea*. First, we found that the profile of the temperature of maximum expression for the thermally-responsive genes differed between species (Figure 3) with a similar number of genes at high temperatures, more genes at intermediate temperatures in *A. carolinensis* and more genes at low temperatures in *A. picea*.

@@

We quantified differences in the expression reaction norms of thermally-responsive genes between species as changes to due shifts in mean ($\Delta M$), slope ($\Delta S$), curvature ($\Delta C$) and wiggle ($\Delta W$). Mean expression (*t*~`r t.mean$parameter`~ = `r t.mean$statistic`, *P* < `r max(t.mean$p.value, 0.001)`) was about `r round(exp(t.mean$estimate))` TPM greater in *A. carolinensis*. Mean slope (*t*~`r t.slope$parameter`~ = `r t.slope$statistic`, *P* < `r max(t.slope$p.value, 0.001)`) was about `r round(exp(t.mean$estimate))` $\Delta$TPM/degree greater in *A. picea*. Curvature (*t*~`r t.curvature$parameter`~ = `r t.curvature$statistic`, *P* < `r round(max(t.curvature$p.value, 0.001), 3)`) and wiggle (*t*~`r t.wiggle$parameter`~ = `r t.wiggle$statistic`, *P* < `r max(t.wiggle$p.value, 0.001)`) did not differ between species. Of the total differences ($\Delta T$) in reaction norms between species for each gene, about ...% (95% CI ... - ...%) was due to shifts in mean and ...% (95% CI ... - ...%) due to shifts in slope.

We predicted that we would observe that genes up-regulated in response to thermal stress would have increased basal expression in the other species. Instead, we found that for *High* genes in *A. picea* (Wilcoxon V = `r w1$statistic`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r w1$conf.int[1:2]`), *Low* genes in *A. carolinensis* (Wilcoxon V = `r w2$statistic`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r w2$conf.int[1:2]`) and *Intermediate* genes in both (Wilcoxon V = `r w3$statistic`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r w3$conf.int[1:2]`), basal expression near optimum temperatures of 19.5C was always higher in *A. carolinensis*.

Sensitivity of thermally-responsive genes is indicated by the thermal range of *Bimodal* expressed genes.

We tested for a transcriptome-wide pattern of thermal sensitivity and tolerance by examining the thermal range for the *Bimodal* and *Intermediate* expressed genes. We found no in thermal sensitivity between species for the *Bimodal* expressed genes (*t*~df=1406~ = 0.31, *P* = 0.76). However, for *Intermediate* genes, *A. carolinensis* had a broader thermal range of stable gene expression (*t*~df=1406~ = 9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64).

We also explored the extent to which the temperature at which thermally-responsive genes are activated differs between species. For genes that are upregulated at *High* temperatures, we found that *A. carolinensis* genes were activated on average 1°C hotter than *A. picea*. In contrast, for genes that are upregulated at *Low* temperaturese we found that...

@@

Thermal tolerance is a key trait for determining the survival of individual organisms, the persistence of populations and the geographic limits of species. While considerable research has focused on a few key genes (e.g. heat shock proteins) involved in thermal stress response [...], little is known about the functional diversity or transcriptome-wide patterns of thermal responsiveness. In this study, we characterized the thermal reactionome of sister ant species, *Aphaenogaster picea* and *Aphaenogaster carolinensis*. We found that nearly 10% of the genes in this species show significant thermal responsiveness.

>>>>

---
title: "Thermal reactionome of a temperate ant species"
author: "John Stanton-Geddes, Andrew Nguyen, Mahesh Vangala, Jim Vincent, Rob Dunn,
  Aaron Ellison, Nate Sanders, Nick Gotelli, Sara Helms-Cahan"
csl: ecology.csl
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    latex_engine: xelatex
  html_document: null
bibliography: ApTranscriptome.bib
---

@@

When ants are exposed to different temperatures, lots of genes change expression. The number of responsive genes suggest that there is greater adaptability to cold than hot thermal stress.

Comparisons between species indicate that the species from the warmer environment tolerates thermal stress by higher baseline expression of protective genes

@@

While characterizing the reactionome is useful for description of the molecular pathways for thermal tolerance, a greater understanding of the process of evolution can be gained using a comparative approach [@harvey1991]. We compared the thermal reactionomes between two parapatric species of *Aphaenogaster*, *A. carolinensis* from North Carolina, USA and *A. picea* from Vermont, USA, to determine the extent to which geographic differentiation and selection has changed the expression patterns of expression for genes involved in thermal tolerance. Specifically, we examined differences between the species in (i) the number of thermally-responsive genes, (ii) the distribution of responsive genes into functional categories, (iii) basal expression near optimal temperatures for responsive genes and (iv) the thermal sensitivity of the transcriptome. *A. picea*, the northern species, experiences less chronic and acute heat stress than *A. carolinensis* [@diamond2013] we predicted that it would show more plasticity of gene expression in response to heat shock, and lower basal levels of expression for heat-tolerant genes.

@@

To describe the overall molecular toolkit for thermal tolerance in *Aphaenogaster*, we performed gene set enrichment analysis (GSEA) on the intersection of each expression category in the two species using the program `topGO` [@alexa2006; @alexa2010] with the `parentChild` algorithm [@grossmann2007] implemented in R [@r_core_team2013]. Briefly, this approach identifies GO terms that are overrepresented in the significant genes relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms. In addition, we used the web-tool DAVID for ....

@@

We identified `r prettyNum(length(which(RxNpval$pval < 0.05)), big.mark = ",", scientific = FALSE)` genes with overall model fits of *P* < 0.05, retaining `r prettyNum(length(which(RxNpval$pval < 0.05)), big.mark = ",", scientific = FALSE)` after adjusting *P* values for false discovery rate. Of these genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` significantly change expression with temperature including `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` that have a species x temperature interaction. The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not thermally-responsive in one species (*NotResp*) differed significantly between the species ($\chi^2$ = `r chi1$statistic`, P < `r max(chi1$p.value, 0.001)`; Table 1). The density function for the temperature of maximum expression (Fig. 1) showed that the majority of responsive genes were in the *Low* category for both species, with *A. picea* having more *Low* genes than *A. carolinensis*. More genes had *Intermediate* expression in *A. carolinensis* than in *A. picea*, while the two species had a similar number of *High* genes (Fig. 1, Table 1).

@@

@@

The reaction norms of thermally-responsive genes between species differed by mean expression (*t*~`r t.mean$parameter`~ = `r t.mean$statistic`, *P* < `r max(t.mean$p.value, 0.001)`) which was about `r round(exp(t.mean$estimate))` TPM greater in *A. carolinensis* than *A picea*, and mean slope (*t*~`r t.slope$parameter`~ = `r t.slope$statistic`, *P* < `r max(t.slope$p.value, 0.001)`), which was about `r round(exp(t.mean$estimate))` $\Delta$TPM/degree greater in *A. picea*. Curvature (*t*~`r t.curvature$parameter`~ = `r t.curvature$statistic`, *P* < `r round(max(t.curvature$p.value, 0.001), 3)`) and wiggle (*t*~`r t.wiggle$parameter`~ = `r t.wiggle$statistic`, *P* < `r max(t.wiggle$p.value, 0.001)`) did not differ between species. Of the total differences ($\Delta T$) in reaction norms between species for each gene, about ...% (95% CI ... - ...%) and ...% (95% CI ... - ...%) were due to shifts in mean and slope, respectively.

We predicted that we would observe that genes up-regulated in response to thermal stress would have increased basal expression in the other species. Instead, we found that for *High* genes 
*A. picea* (Wilcoxon V = `r w1$statistic`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r w1$conf.int[1:2]`), *Low* genes in *A. carolinensis* (Wilcoxon V = `r w2$statistic`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r w2$conf.int[1:2]`) and *Intermediate* genes in both (Wilcoxon V = `r w3$statistic`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r w3$conf.int[1:2]`), basal expression near optimum temperatures of 19.5C was always higher in *A. carolinensis*.

Overall greater thermal tolerance of one species would be indicated by finding that gene expression is stable across a greater portion of the thermal gradient. For *Bimodal* expressed genes, we found no difference (*t*~df=1406~ = 0.31, *P* = 0.76) between species in the variance of the reaction norms for thermally-responsive genes. In contrast, *Intermediate* genes from *A. carolinensis* had a broader thermal range of stable gene expression than those from *A. picea* (*t*~df=1406~ = 9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64). In addition, the species differed in the temperatures at which thermally-responsive genes were activated. For genes with increased expression at *High* temperatures, we found that *A. carolinensis* genes were activated on average 1°C hotter than *A. picea* genes (Fig. 2). Similarly, for genes with increased expression at *Low* temperatures, *A. carolinensis* genes were activated on average 1°C hotter than *A. picea* genes (Fig. 2). Thus, while the activation temperatures differed, the size of the window between activation of *Low* and *High* genes remained the same between species, consistent with the results for *Bimodal* genes above.

@@

Thermal tolerance is a key trait for determining the survival of individual organisms, the persistence of populations and the geographic limits of species. While considerable research has focused on a few key genes (e.g. heat shock proteins) involved in thermal stress response [...], little is known about the functional diversity or transcriptome-wide patterns of thermal responsiveness. In this study, we characterized the thermal reactionome of parapatric ant species, *Aphaenogaster picea* and *Aphaenogaster carolinensis* found in colder and warmer environments, respectively. We found that nearly 10% of the genes in this species show significant thermal responsiveness.
